---
title: 'ERP classification by sub'
author: Andrew H. Farkas
date: "`r format(Sys.time(), '%e %B, %Y')`"
output:
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(fdapace)
library(fda)
library(e1071) # support vector machine
#library(bestglm)
#library(caret)
#library(glmnet)

# Make standard error function
se <- function(vec, na.rm = FALSE) {
  sd(vec, na.rm)/sqrt(n())
}

#options("scipen"=100, "digits"=4)
```

# load data sub by pic

```{r}
load(file = "C:/Users/andre/Documents/R/data/all_mas.Rdata")
```

# tidy data

<!-- test to see if data extraction worked -->
<!-- ```{r} -->
<!-- all_mas %>% -->
<!--   filter(electrode == 31) %>% -->
<!--   group_by(PictureCat) %>% -->
<!--   summarise_all(mean) %>% -->
<!--   pivot_longer(cols = starts_with("ms"), -->
<!--                names_to = "time_ms", -->
<!--                names_prefix = "ms", -->
<!--                values_to = "voltage") -> hold -->

<!-- hold$time_ms <- hold$time_ms %>% as.numeric() -->
<!-- hold$CatNum <- hold$CatNum %>% as.factor() -->

<!-- hold %>% -->
<!--   ggplot(aes(x = time_ms, y = voltage, color = PictureCat)) + -->
<!--   geom_line() -->

<!-- rm(hold) -->
<!-- ``` -->

```{r}
all_mas <- all_mas %>% filter(electrode != 65)

all_mas$Picture.number <- all_mas$Picture.number %>% as.factor()
all_mas$PictureCat <- all_mas$PictureCat %>% as.factor()
all_mas$CatNum <- all_mas$CatNum %>% as.factor()
all_mas$subid <- all_mas$subid %>% as.factor()
all_mas$electrode <- all_mas$electrode %>% as.factor()

all_mas %>% head()
```

# Predict 2 categories

```{r}
prediction_categories <- c("Erotica", "Bodies")
filtered_mas <- all_mas %>% filter(PictureCat %in% prediction_categories)
```


for sub i in sub id
for each cat
assign erps to one of 3 folds
for each fold
assign fold to test data
train support vector on fPCA train
find fPCA for test based on train
predict test
save correct percentage

```{r}
#for testing
sub_index <- 1
cat_index <- 1
cat_index <- 2
channel <- 1
fold_index <- 1

sub_ids <- filtered_mas$subid %>% unique()
number_of_subs <- sub_ids %>% length()

hold_names <- names(filtered_mas)[8:ncol(filtered_mas)]

time_ms <- substr(hold_names, start = 3, stop = nchar(hold_names)) %>% 
  as.numeric()

rm(hold_names)

range_time <- range(time_ms)

breaks <- c(-100, 0, 
            seq(5,100,5), 
            seq(110, 300, 10), 
            seq(350, 900, 50),
            seq(1000, 2000, 100))


bspline_basis <- create.bspline.basis(rangeval = range_time,
                                      # nbasis = 42,
                                      norder = 3,
                                      breaks = breaks)

b_vals <- eval.basis(time_ms, bspline_basis)
X_basis <- b_vals

set.seed(0)

for (sub_index in 1:number_of_subs) {
  current_sub <- sub_ids[sub_index]
  
  current_dat <- filtered_mas %>% filter(subid == current_sub)
  
  channels <- current_dat$electrode %>% unique() %>% length()
  
  current_dat$fold_id <- NA
  
  categories <- current_dat$PictureCat %>% unique()
  number_of_cats <- categories %>% length()
  
  for (cat_index in 1:number_of_cats) {
    current_cat <- categories[cat_index]
    
    fold_key <- current_dat %>% filter(PictureCat == current_cat) %>% 
      dplyr::select(Picture.number) %>% unique() %>% 
      mutate(folds = sample(rep(1:3, length.out = n())))
    
    fold_1 <- fold_key[fold_key$folds ==1,]$Picture.number
    fold_2 <- fold_key[fold_key$folds ==2,]$Picture.number
    fold_3 <- fold_key[fold_key$folds ==3,]$Picture.number
    
    current_dat[current_dat$Picture.number %in% fold_1,]$fold_id <- 1
    current_dat[current_dat$Picture.number %in% fold_2,]$fold_id <- 2
    current_dat[current_dat$Picture.number %in% fold_3,]$fold_id <- 3
  }
  
  for (fold_index in 1:3) {
    
    test_scores <- current_dat[,1:7]
    test_scores$fold_id <- current_dat$fold_id
    test_scores$scores <- NA
    
    for (channel in 1:channels){
      
      current_dat_one_chan <- current_dat %>% 
        filter(electrode == channel)
      
      just_volts_over_time <- 
        current_dat_one_chan[, 8:(ncol(current_dat_one_chan)-1)] #minus one to exclude fold_id
      
      #Preallocate
      smooth_curves <- array(0, dim(just_volts_over_time))
      
      rows <- nrow(just_volts_over_time)
      
      for(row_index in 1:rows){
        y_dat <- unlist(just_volts_over_time[row_index,])
        fit   <- lm(y_dat ~ 0 + X_basis)
        smooth_curves[row_index,] <- fit$fitted.values
      }
      
      mean_hat <- colMeans(smooth_curves[current_dat_one_chan$fold_id != fold_index,])
      
      smooth_cov <- cov(smooth_curves[current_dat_one_chan$fold_id != fold_index,])
      
      svd_result0 <- eigen(smooth_cov, symmetric = TRUE)
      e_vectors <- svd_result0$vectors[,svd_result0$values > 0]
      e_values <- svd_result0$values[svd_result0$values > 0]
      
      e_fns0 <- e_vectors*sqrt(ncol(just_volts_over_time))
      e_vals0 <- e_values/ncol(just_volts_over_time)
      
      #Maybe move out of loop
      n_pc <- 1
      
      # truncate estimated eigen components
      e_fns <- e_fns0[,1:n_pc]
      e_vals <- e_vals0[1:n_pc]
      
      #concerned about this demeaning when my data is 
      erp_demeaned <- just_volts_over_time -
        t(matrix(rep(mean_hat,
                     rows),
                 nrow=length(mean_hat)))
      
      erp_scores <- matrix(NA, nrow = rows, ncol = n_pc)
      
      
      for(row_index in 1:rows){
        erp_scores[row_index,] <- 
          colMeans(matrix(rep(as.numeric(erp_demeaned[row_index,]), 
                              n_pc), nrow=ncol(just_volts_over_time)) * e_fns)
      }
      
      test_scores[test_scores$electrode == channel,]$scores <-
        erp_scores
      
    }
    
    dat_train <- test_scores %>% 
      filter(fold_id != fold_index)
    
    dat_test <- test_scores %>% 
      filter(fold_id == fold_index)
    
    dat_train %>% 
      pivot_wider(names_from = electrode, values_from = scores) ->
      dat_train_wide
    
    names(dat_train_wide)[8:71] <- paste0("channel_",names(dat_train_wide)[8:71])
    
    dat_test %>% 
      pivot_wider(names_from = electrode, values_from = scores) ->
      dat_test_wide
    
    names(dat_test_wide)[8:71] <- paste0("channel_",names(dat_test_wide)[8:71])
    
    train_for <- as.formula(paste(colnames(dat_train_wide)[3], 
                                  "~",
                            paste(colnames(dat_train_wide)[c(8:71)],
                                  collapse = "+"),
                                  sep = ""))
    
    model_svm = svm(train_for, data = dat_train_wide)
    
    pred <- predict(model_svm, dat_train_wide)
    
    # Model accuracy
    mean(pred==dat_train_wide$PictureCat)
    
    pred <- predict(model_svm, dat_test_wide)
    mean(pred==dat_test_wide$PictureCat)
    
  }
  
}


```

```{r}
rep(1:3, length.out = 5) %>% sample()

train_dat[4,8:ncol(train_dat)] %>% unlist() %>% plot()
```


in algorithm

# make 3 equally sized sets of ERPs per sub

# get fpca scores for training and test

# train and test 

# do for other folds

for number of subjects

split ERPs in to 3 equally sized sets
  
for number of sets

pull one set for testing
    
train on other sets
    
test on final set
    
save correct %
    
