---
title: 'ERP classification by sub'
author: Andrew H. Farkas
date: "`r format(Sys.time(), '%e %B, %Y')`"
output:
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(fdapace)
library(fda)
#library(bestglm)
#library(caret)
#library(glmnet)

# Make standard error function
se <- function(vec, na.rm = FALSE) {
  sd(vec, na.rm)/sqrt(n())
}

#options("scipen"=100, "digits"=4)
```

# load data sub by pic

```{r}
load(file = "C:/Users/andre/Documents/R/data/all_mas.Rdata")
```

# tidy data

<!-- test to see if data extraction worked -->
<!-- ```{r} -->
<!-- all_mas %>% -->
<!--   filter(electrode == 31) %>% -->
<!--   group_by(PictureCat) %>% -->
<!--   summarise_all(mean) %>% -->
<!--   pivot_longer(cols = starts_with("ms"), -->
<!--                names_to = "time_ms", -->
<!--                names_prefix = "ms", -->
<!--                values_to = "voltage") -> hold -->

<!-- hold$time_ms <- hold$time_ms %>% as.numeric() -->
<!-- hold$CatNum <- hold$CatNum %>% as.factor() -->

<!-- hold %>% -->
<!--   ggplot(aes(x = time_ms, y = voltage, color = PictureCat)) + -->
<!--   geom_line() -->

<!-- rm(hold) -->
<!-- ``` -->

```{r}
all_mas$Picture.number <- all_mas$Picture.number %>% as.factor()
all_mas$PictureCat <- all_mas$PictureCat %>% as.factor()
all_mas$CatNum <- all_mas$CatNum %>% as.factor()
all_mas$subid <- all_mas$subid %>% as.factor()
all_mas$electrode <- all_mas$electrode %>% as.factor()

all_mas %>% head()
```

# Predict 2 categories

```{r}
prediction_categories <- c("Erotica", "Bodies")
filtered_mas <- all_mas %>% filter(PictureCat %in% prediction_categories)
```


for sub i in sub id
for each cat
count number of ERPs per cat
assign erps to one of 3 folds
for each fold
assign fold to test data
train support vector on fPCA train
find fPCA for test based on train
predict test
save correct percentage

```{r}
#for testing
sub_index <- 1

sub_ids <- filtered_mas$subid %>% unique()
number_of_subs <- sub_ids %>% length()



for (sub_index in 1:number_of_subs) {
  current_sub <- sub_ids[sub_index]
  
  current_dat <- filtered_mas %>% filter(subid == current_sub)
  
  categories <- current_dat$PictureCat %>% unique()
  number_of_cats <- categories %>% length()
  
  for (cat_index in 1:number_of_cats) {
    
    
    
  }
  
}
```



in algorithm

# make 3 equally sized sets of ERPs per sub

# get fpca scores for training and test

# train and test 

# do for other folds

for number of subjects

split ERPs in to 3 equally sized sets
  
for number of sets

pull one set for testing
    
train on other sets
    
test on final set
    
save correct %
    
